A Partitioning Algorithm for Maximum Common (Connected) Subgraphs\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/K503058/1 and EP/M508056/1]}

### 2017-02-07 paper/paper.tex
Added titles of some other sections in paper

\newcommand{\examplexG}[5] {
    \begin{minipage}{.3\textwidth}
    \tikz {
        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1cm,
                clockwise=5] {
                    1[label=90:#1],2[label=0:#2],3[label=0:#3],4[label=180:#4],5[label=180:#5];
            1--4; 1--5; 2--3; 2--5; 3--5;
        };
    }
    \end{minipage}
}
\newcommand{\examplexH}[6] {
    \begin{minipage}{.3\textwidth}
    \tikz {
        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1.1cm,
                clockwise=6, phase=60] {
                    a[label=0:#1],b[label=0:#2],c[label=0:#3],d[label=180:#4],e[label=180:#5],f[label=180:#6];
            a--b; a--c; a--e; b--d; b--f; c--d; c--e; c--f; d--f; e--f;
        };
    }
    \end{minipage}
}

@@

Given graphs $G=(V^G, E^G)$ and $H=(V^H, E^H)$, the \emph{Maximum Common
Subgraph (MCS)} problem is to find a graph with as many vertices as possible
that is isomorphic to induced subgraphs of both $G$ and $H$. Equivalently, MCS
is the problem of finding a maximum-cardinality matching $M = \{(v_1, w_1),
\dots, (v_{|M|}, w_{|M|})\}$, such that for each pair $i,j$ ($1 \leq i \leq j
\leq |M|$), $v_i$ and $v_j$ are adjacent in $G$ if and only if $w_i$ and $w_j$
are adjacent in $H$.

The most successful existing approaches to optimally solving MCS are
reformulation to the maximum clique problem and constraint programming
\cite{DBLP:conf/cp/McCreeshNPS16}. In this paper, we introduce a new branch and bound
algorithm for MCS which does not require a domain to be explicity maintained in
memory for each vertex in $V^G$. The algorithm maintains labellings of the
unmatched vertices in $V^G$ and $V^H$ such that a pair $(v, w)$ with $v \in V^G$
and $w \in V^H$ can be added to the matching if and only if $v$ and $w$ have the
same label.

@@

\begin{figure}[ht]
\centering
    \examplexG{}{}{}{}{}
    \examplexH{}{}{}{}{}{}
\caption{Step 1}
\label{fig:alg1}
\end{figure}

The algorithm proceeds by building up a mapping $M$, starting with the empty
mapping $\{\}$. Begin by (arbitrarily) mapping vertex $1$ in $G$ to vertex $a$
in $H$, giving $M=\{1a\}$.  Label each unmatched vertex according to whether it
is adjacent to $1$ (in $G$) or $a$ (in $H$), as shown in Figure~\ref{fig:alg2}.
Adjacent vertices have label 1; non-adjacent vertices have label 0.  It is
straightforward to see that we can extend $M$ with a vertex mapping $vw$, with
$v \in V^G$ and $w \in V^H$, if and only if $v$ and $w$ have the same label.  This
property is the algorithm's key invariant.

@@

Next, extend the mapping by pairing a vertex in $G$ with a vertex in $H$ of the
same label; we will choose to map vertex $2$ to vertex $d$, giving $M=\{1a,
2d\}$ (Figure~\ref{fig:alg3}).  Each unmatched vertex $v \in V^G$ is labelled
with a two-character bit string, indicating whether $v$ is adjacent to each of
the two matched vertices in $V^G$ (vertices $1$ and $2$).  For example, vertex
$3$ is labelled $01$, indicating that it is not adjacent to $1$ but is adjacent
to $2$.  Labels are given to vertices in $V^H$ in a similar fashion, showing
whether each vertex is adjacent to $a$ and $d$.

@@

The algorithm proceeds by backtracking search. When we have explored the full
search space of matchings containing $\{1a, 2d\}$, we try reassigning $2$ to
$f$.  Since $d$ and $f$ are the only vertices to which $2$ can be matched given
the decision to match $1$ to $a$, we lastly explore the possibility that $2$ is
left unmatched, by giving $2$ the label $\bot$ and selecting another vertex in
$V^G$ to assign.

@@

@@

>>>>

\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\E}{E}

\newcommand{\examplexG}[5] {
    \begin{minipage}{.2\textwidth}
    \tikz {
        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1cm,
                clockwise=5] {
                    1[label=90:#1],2[label=0:#2],3[label=0:#3],4[label=180:#4],5[label=180:#5];
            1--4; 1--5; 2--3; 2--5; 3--5;
        };
    }
    \end{minipage}
}
\newcommand{\examplexH}[6] {
    \begin{minipage}{.2\textwidth}
    \tikz {
        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1.1cm,
                clockwise=6, phase=60] {
                    a[label=0:#1],b[label=0:#2],c[label=0:#3],d[label=180:#4],e[label=180:#5],f[label=180:#6];
            a--b; a--c; a--e; b--d; b--f; c--d; c--e; c--f; d--f; e--f;
        };
    }
    \end{minipage}
}

@@

Given graphs $G=(\V(G), \E(G))$ and $H=(\V(H), \E(H))$, the \emph{Maximum Common
Subgraph (MCS)} problem is to find a graph with as many vertices as possible
that is isomorphic to induced subgraphs of both $G$ and $H$. Equivalently, MCS
is the problem of finding a maximum-cardinality set of pairs $M = \{(v_1, w_1),
\dots, (v_{|M|}, w_{|M|})\}$, with $v_i \in V(G)$ and $v_i \in V(H)$ for all $i$,
such that for each pair $i,j$ ($1 \leq i < j
\leq |M|$), $v_i$ and $v_j$ are adjacent in $G$ if and only if $w_i$ and $w_j$
are adjacent in $H$.

The most successful existing approaches to optimally solving MCS are
reformulation to the maximum clique problem and constraint programming
\cite{DBLP:conf/cp/McCreeshNPS16}. In this paper, we introduce a new branch and bound
algorithm for MCS which does not require a domain to be explicity maintained in
memory for each vertex in $\V(G)$. The algorithm maintains labellings of the
unmatched vertices in $\V(G)$ and $\V(H)$ such that a pair $(v, w)$ with $v \in \V(G)$
and $w \in \V(H)$ can be added to the matching if and only if $v$ and $w$ have the
same label.

@@

\begin{figure}[ht]
\centering
    \examplexG{}{}{}{}{}
    \examplexH{}{}{}{}{}{}
\caption{Graphs $G$ and $H$}
\label{fig:alg1}
\end{figure}

The algorithm proceeds by building up a mapping $M$, starting with the empty
mapping $\{\}$. Begin by (arbitrarily) mapping vertex $1$ in $G$ to vertex $a$
in $H$, giving $M=\{1a\}$.  Label each unmatched vertex according to whether it
is adjacent to $1$ (in $G$) or $a$ (in $H$), as shown in Figure~\ref{fig:alg2}.
Adjacent vertices have label 1; non-adjacent vertices have label 0.  It is
straightforward to see that we can extend $M$ with a vertex mapping $vw$, with
$v \in \V(G)$ and $w \in \V(H)$, if and only if $v$ and $w$ have the same label.  This
property is the algorithm's key invariant.

@@

Next, extend the mapping by pairing a vertex in $G$ with a vertex in $H$ of the
same label; we will choose to map vertex $2$ to vertex $d$, giving $M=\{1a,
2d\}$ (Figure~\ref{fig:alg3}).  Each unmatched vertex $v \in \V(G)$ is labelled
with a two-character bit string, indicating whether $v$ is adjacent to each of
the two matched vertices in $\V(G)$ (vertices $1$ and $2$).  For example, vertex
$3$ is labelled $01$, indicating that it is not adjacent to $1$ but is adjacent
to $2$.  Labels are given to vertices in $\V(H)$ in a similar fashion, showing
whether each vertex is adjacent to $a$ and $d$.

@@

The algorithm proceeds by backtracking search. When we have explored the full
search space of matchings containing $\{1a, 2d\}$, we try reassigning $2$ to
$f$.  Since $d$ and $f$ are the only vertices to which $2$ can be matched given
the decision to match $1$ to $a$, we lastly explore the possibility that $2$ is
left unmatched, by giving $2$ the label $\bot$ and selecting another vertex in
$\V(G)$ to assign.

@@

\section{Heuristics}

\section{Extensions}

So far, we have considered only unlabelled, undirected graphs. In this section, we
describe some straightforward changes to the algorithm that relax these restrictions.

For the case of unlabelled, directed graphs, each adjacency-matrix entry for a vertex
pair $(u,v)$ in graph $G$ or $H$ takes the value 0 if the vertices are not adjacent,
1 if there is an edge $u \rightarrow v$ but no edge $v \rightarrow u$, 2 if there is a
edge in the opposite direction, and 3 if there are edges in both directions. In lines
14-19 of the algorithm .......

Labelled, undirected .......

Labelled, directed ......

Connected ........

\section{Future Work}

Instead of branching on vertices in $\V(G)$, it would be equally valid to branch on
a vertex in $\V(H)$ in lines 9-23 of the algorithm, since our data structure treats the
two graphs symmetrically. More interestingly, we could choose which graph to branch on
at each search node using some heuristic (perhaps choosing based on whether the $P$ or
$T$ set is smaller).

We could create a hybrid of Lilybank and clique (for example, using Lilybank near the
top of the search tree, but switching to a clique encoding when fewer than some threshold
number of vertices remain to be selected). This might give us most of the benefits of
the clique encoding for labelled graphs, while avoiding the high memory cost and colouring
time of encoding the full instance.

@@

\section{Conclusion}
